Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-03-17 ltx/BrainIDE.tex
Some minor changes to the derivations and some questions

In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.
The model relates the average number of action potentials $g(r,t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(r,t)$. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by

@@

@@

@@

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function

@@

	f\left( v_t\left( r' \right) \right) = \frac{\nu _0\left(r\right)}{1 + \exp \left( \beta \left( v_0\left( r \right) - v\left(r',\bar{t} \right) \right) \right)}.

@@

\section{Generating Data}
To generate iEEG/LFP data we proceeded with the assumption of spatial homogeneity for the connectivity and activation function parameters, implying $w(r,r') = w\left(|r-r'|\right)$, $\nu_0(r) = \nu_0$ and $v_0(r) = v_0$. In addition we assume the synaptic transmission delay is negligible such that $\bar{t}=t$. The model was discretized in time  (see Appendix~\ref{Time Discretization}) and a stochastic term was added to drive the system and account for uncertainty in the model giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I_t\left(r\right) + e_t\left(r\right)
\end{equation}
where $\delta$ is the time step, $\lambda = 1-\alpha\delta$ and $e_t(r) \sim \mathcal{GP}(0,\Sigma)$. The IDE formulation in Eq.~\ref{DiscreteTimeModel} was solved using a numerical method described in Appendix~\ref{Space Discretization}. To generate iEEG/LFP data we used the output function 
\begin{equation}
	y_t\left(r_n\right) = \int_{\Omega}{m\left(|r_n-r'|\right)v_t\left(r\right)dr'}
\end{equation}
where $r_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors. The output kernel $m(|r-r'|)$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(|r-r'|\right) = \exp{\left(-\frac{\left(r_x - r_{x_n}\right)^2}{\left(2\sigma_m\right)^2}-\frac{\left(r_y - r_{y_n}\right)^2}{\left(2\sigma_m\right)^2}\right)}.
\end{equation}

@@

\section{Derivation of Finite Dimensional State-Space Model}
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
	v_t\left(r\right) \approx \phi ^T\left(r\right) x_t
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $x_t$. As described in Eq.~\ref{Connectivity Kernel} the connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	\tilde{w}( |r-r'| ) = \theta^T\psi\left(|r-r'|\right)
\end{equation}
Will will assume we know the parametric form of the connectivity basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get
\begin{equation}\label{reduced continuous model}
	\phi^T\left( r \right)x_{t+1} = \delta\theta^T\int_\Omega  {\psi \left(|r-r'|\right)f\left( \phi ^T\left(r'\right)x_t \right)dr'}  - \lambda\phi^T\left(r\right)x_t + \delta I_t\left(r\right) + e_t\left(r\right).
\end{equation}
By defining the term
\begin{equation}\label{DefGamma}
	\Gamma  = \int_\Omega  {\phi \left(r\right)\phi ^T\left(r\right)dr}
\end{equation}
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model
\begin{eqnarray}\label{Homogeneous SS Model}
	x_{t+1} &=& \delta\Gamma^{ - 1}\int_\Omega {\phi\left(r\right) \int_\Omega {f\left(\phi^T\left(r'\right)x_t\right)\psi^T \left(|r-r'|\right)dr'} dr} \theta - \lambda x_t + \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} + \int_\Omega{\phi\left(r\right)\tilde{e}_t\left(r\right)dr} \\
	&=&\delta q(x_t) \theta -\lambda x_t + + \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} + \int_\Omega{\phi(r)\tilde{e}_t(r)dr}
\end{eqnarray}
where $\tilde{e}_t\left(r\right) \sim \mathcal{N}\left(0,R' \Sigma R\right)$ and $R$ is the Cholesky decomposition of $\Gamma^{-1}$. The observation equation of the reduced model is
\begin{eqnarray}\label{ObservationEquation}
    y_t( r_n ) &=& \int_\Omega  m\left(|r_n-r|\right)\phi^T\left(r\right) dr x_t  + \varepsilon_t ( r_n ) \\
	&=& C x_t + \varepsilon_t ( r_n ).
\end{eqnarray}

@@

\section{Discrete Time Model}\label{Time Discretization}
Using a one-step Euler method we perform the time discretization by
To form the IDE neural field model a time discretization must be perform. We used a one-step Euler method where Eq.~\ref{FinalForm1} can be approximated by
\begin{equation}\label{Euler Approximation}	
\frac{v\left( r,t+\delta \right) - v\left( r,t\right)}{\delta} =   -\alpha v\left( r,t \right) + \int_\Omega  {w\left( |r-r'| \right)f\left( {v\left( r',t \right)} \right)dr'} + I\left(r,t\right).
\end{equation}
For clarity, we shall index time points in the discrete time form of the model using the subscript $t$ and the next time point as $t+1$. Rearranging Eq.~\ref{Euler Approximation} we get
\begin{eqnarray}\label{Euler Approximation}	
v_{t+1}\left( r\right) &=& v_t\left( r\right) -\delta \alpha v_t\left( r\right) + \delta \int_\Omega  {w\left( |r-r'| \right)f\left( {v_t\left( r'\right)} \right)dr'} + \delta I_t\left(r\right) \\
&=& \lambda v_t(r) + \delta \int_\Omega  {w\left( |r-r'| \right)f\left( {v_t\left( r'\right)} \right)dr'} + \delta I_t\left(r\right)
\end{eqnarray}
The discrete time form of the model is
\begin{equation}\label{Discrete Time Model1}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I\left(r,t\right).
\end{equation}
where $\lambda = 1 - \delta \alpha$.

\section{Numerical Solution to IDE Model}\label{Space Discretization}
To solve the PDE we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v_{t+1}\left(r_{ij}\right) =  \lambda v_t(r_{ij}) + \delta \Delta^2 \sum_{i=1}^{i=I}{\sum_{j=1}^{j=J}{\tilde{w}\left( |r_{ij}-r_{ij}'| \right)f\left( v_t\left( r_{ij}'\right) \right)} } + \delta I_t\left(r\right) + \tilde{e}_t(r_{i,j})
\end{equation}
where $\tilde{e}_t(r_{i,j}) \sim \mathcal{N}\left(0,\Sigma\right)$ and $\tilde{w}(|r_{ij}-r_{ij}'|)$ discretized connectivity kernel described by
\begin{equation}\label{Discrete Connectivity Kernel}
	\tilde{w}(r_{ij},r_{ij}') = \theta^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\sigma)^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\sigma)^2}}.
\end{equation}
The observation equation and kernel were discretized in the same way by
\begin{equation}\label{Discrete Observation Equation}
    y_k\left(r_{i_n j_n}\right) = \Delta^2\sum_{i=1}^{i=I}\sum_{j=1}^{j=J} \tilde{m}( |r_{i_n j_n}-r_{ij}'| ) v_k\left(r_{ij}'\right)  + \varepsilon_k( r_{i_n j_n})
\end{equation}
where
\begin{equation}\label{Discrete Observation Kernel}
	\tilde{m}(|r_{ij}-r_{ij}'|) = e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\sigma)^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\sigma)^2}}.
\end{equation}
We have used periodic boundary conditions to alleviate problems associated with edge effects.

@@

Starting from Eq.~\ref{reduced continuous model} we crossing multiply and integrate over the field to get

@@

Substituting Eq.~\ref{DefGamma} into Eq.~\ref{Poo} and cross-multiplying by $\Gamma^{-1}$ gives the state-space model

@@

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{Colored noise}
We have
\begin{equation}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( \bf{r'} \right)} \right)d{\bf{r'}}} + \lambda v_t\left( {\bf{r}} \right) + T_se_t\left( {\mathbf{r}} \right).
\end{equation}
where the disturbance $e_t(\mathbf r)$ is a zero-mean normally distributed noise process, spatially coloured but temporally independent, with covariance

>>>>

In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.}
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.}
\end{todo}
The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology. The model relates the average number of action potentials $g(r,t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(r,t)$. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by

@@

\begin{todo}
	{Is this equation correct? Is it a convolution in time?}
\end{todo}

@@

\begin{todo}
	{Do we really need this representation with $\bar t$. It makes convolution look wrong.}
\end{todo}

@@

\begin{todo}
	{Parham ... how are you writing this?}
\end{todo}The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function

@@

	f\left( v\left( r', \bar t \right) \right) = \frac{\nu _0\left(r\right)}{1 + \exp \left( \beta \left( v_0\left( r \right) - v\left(r',\bar{t} \right) \right) \right)}.

@@

\section{Generating Data}
To generate iEEG/LFP data we proceeded with the assumption of spatial homogeneity for the connectivity and activation function parameters, implying $w(r,r') = w\left(|r-r'|\right)$, $\nu_0(r) = \nu_0$ and $v_0(r) = v_0$. In addition we assume the synaptic transmission delay is negligible such that $\bar{t}=t$. The model was discretized in time  (see Appendix~\ref{Time Discretization}) and a stochastic term was added to drive the system and account for uncertainty in the model giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I_t\left(r\right) + e_t\left(r\right)
\end{equation}
where $\delta$ is the time step, $\lambda = 1-\alpha\delta$ and $e_t(r) \sim \mathcal{GP}(0,\Sigma)$. The IDE formulation in Eq.~\ref{DiscreteTimeModel} was solved using a numerical method described in Appendix~\ref{Space Discretization}. To generate iEEG/LFP data we used the output function 
\begin{equation}
	y_t\left(r_n\right) = \int_{\Omega}{m\left(|r_n-r'|\right)v_t\left(r'\right)dr'}
\end{equation}
where $r_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors. The output kernel $m(|r-r'|)$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(|r-r'|\right) = \exp{\left(-\frac{\left(r_x - r_{x_n}\right)^2}{\left(2\sigma_m\right)^2}-\frac{\left(r_y - r_{y_n}\right)^2}{\left(2\sigma_m\right)^2}\right)}.
\end{equation}

@@

\section{Derivation of Finite Dimensional State-Space Model}
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
	v_t\left(r\right) \approx \phi ^T\left(r\right) x_t
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $x_t$. As described in Eq.~\ref{Connectivity Kernel} the connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	\tilde{w}( |r-r'| ) = \theta^T\psi\left(|r-r'|\right)
\end{equation}
We will assume we know the parametric form of the connectivity basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get
\begin{equation}\label{reduced continuous model}
	\phi^T\left( r \right)x_{t+1} = \delta\theta^T\int_\Omega  {\psi \left(|r-r'|\right)f\left( \phi ^T\left(r'\right)x_t \right)dr'}  - \lambda\phi^T\left(r\right)x_t + \delta I_t\left(r\right) + e_t\left(r\right).
\end{equation}
By defining the term
\begin{equation}\label{DefGamma}
	\Gamma  = \int_\Omega  {\phi \left(r\right)\phi ^T\left(r\right)dr}
\end{equation}
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model
\begin{eqnarray}\label{Homogeneous SS Model}
	x_{t+1} &=& \delta\Gamma^{ - 1}\int_\Omega {\phi\left(r\right) \int_\Omega {f\left(\phi^T\left(r'\right)x_t\right)\psi^T \left(|r-r'|\right)dr'} dr} \theta - \lambda x_t + \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} + \int_\Omega{\phi\left(r\right)\tilde{e}_t\left(r\right)dr} \\
	&=&\delta q(x_t) \theta -\lambda x_t + + \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} + \int_\Omega{\phi(r)\tilde{e}_t(r)dr}
\end{eqnarray}
\begin{todo}{Define q}\end{todo}
where $\tilde{e}_t\left(r\right) \sim \mathcal{N}\left(0,R' \Sigma R\right)$ and $R$ is the Cholesky decomposition of $\Gamma^{-1}$. The observation equation of the reduced model is
\begin{eqnarray}\label{ObservationEquation}
    y_t( r_n ) &=& \int_\Omega  m\left(|r_n-r|\right)\phi^T\left(r\right) dr x_t  + \varepsilon_t ( r_n ) \\
	&=& C x_t + \varepsilon_t ( r_n ).
\end{eqnarray}
\begin{todo}{Define C}\end{todo}

@@

\section{Discrete Time Model}\label{Time Discretization}
To form the IDE neural field model a time discretization must be perform. We used a one-step Euler method where Eq.~\ref{FinalForm1} can be approximated by
\begin{equation}\label{Euler Approximation}	
\frac{v\left( r,t+\delta \right) - v\left( r,t\right)}{\delta} =   -\alpha v\left( r,t \right) + \int_\Omega  {w\left( |r-r'| \right)f\left( {v\left( r',t \right)} \right)dr'} + I\left(r,t\right).
\end{equation}
For clarity, we shall index time points in the discrete time form of the model using the subscript $t$ and the next time point as $t+1$. Rearranging Eq.~\ref{Euler Approximation} we get
\begin{eqnarray}\label{Euler Approximation}	
v_{t+1}\left( r\right) &=& v_t\left( r\right) -\delta \alpha v_t\left( r\right) + \delta \int_\Omega  {w\left( |r-r'| \right)f\left( {v_t\left( r'\right)} \right)dr'} + \delta I_t\left(r\right)
\end{eqnarray}
The discrete time form of the model is
\begin{equation}\label{Discrete Time Model1}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I_t\left(r\right).
\end{equation}
where $\lambda = 1 - \delta \alpha$.

\section{Numerical Solution to IDE Model}\label{Space Discretization}
To solve the convolution integral in \ref{Discrete Time Model1} we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v_{t+1}\left(r_{ij}\right) =  \lambda v_t(r_{ij}) + \delta \Delta^2 \sum_{i=1}^{i=I}{\sum_{j=1}^{j=J}{\tilde{w}\left( |r_{ij}-r_{ij}'| \right)f\left( v_t\left( r_{ij}'\right) \right)} } + \delta I_t\left(r\right) + \tilde{e}_t(r_{i,j})
\end{equation}
where $\tilde{e}_t(r_{i,j}) \sim \mathcal{N}\left(0,\Sigma\right)$ and $\tilde{w}(|r_{ij}-r_{ij}'|)$ discretized connectivity kernel described by
\begin{equation}\label{Discrete Connectivity Kernel}
	\tilde{w}(r_{ij},r_{ij}') = \theta^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\sigma)^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\sigma)^2}}.
\end{equation}
The observation equation and kernel were discretized in the same way by
\begin{equation}\label{Discrete Observation Equation}
    y_k\left(r_{i_n j_n}\right) = \Delta^2\sum_{i=1}^{i=I}\sum_{j=1}^{j=J} \tilde{m}( |r_{i_n j_n}-r_{ij}'| ) v_k\left(r_{ij}'\right)  + \varepsilon_k( r_{i_n j_n})
\end{equation}
where
\begin{equation}\label{Discrete Observation Kernel}
	\tilde{m}(|r_{ij}-r_{ij}'|) = e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\sigma)^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\sigma)^2}}.
\end{equation}
We have used periodic boundary conditions to alleviate problems associated with edge effects.

@@

Starting from Eq.~\ref{reduced continuous model} we multiply throughout by $\phi(r)$ and integrate over the spatial domain $\Omega$ to get

@@

Substituting Eq.~\ref{DefGamma} into Eq.~\ref{Poo} and cross-multiplying by $\Gamma^{-1}$ gives 

@@

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{Colored noise}
We have
\begin{equation}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( \bf{r'} \right)} \right)d{\bf{r'}}} + \lambda v_t\left( {\bf{r}} \right) + T_se_t\left( {\mathbf{r}} \right).
\end{equation}
where the disturbance $e_t(\mathbf r)$ is a zero-mean normally distributed noise process, spatially coloured but temporally white, with covariance
